Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation 

### 2025-05-06 manuscript/manuscript.qmd
finished data wrangling

@@

### Limitations of Existing Approaches
Regrettably, the development of traditional ocean color predictive models relies on a fundamental statistical error that plagues most of observational science today; that of conflating sampling probability, with inferential probability([@jaynes2003probability, @DeScheemaekere2011].) Consider a dataset $D$ that consists of input-output pairs - e.g. Remote sensing reflectance (Rrs) and  chlorophyll-*a* concentration $Chl_a$ - and a model $M$, such as OCx, hypothesized to describe the statistical association between them. The sampling probability $p(D|M)$ represents the probability of observing the data $D$ if model $M$ were "true". Standard model fitting maximizes this quantity (i.e. the likelihood) by tuning the parameters of $M$.

The unspoken assumption is that the model that best fits the data also best represents the underlying process. This constitutes an epistemic fallacy; substituting $p(D|M)$ for $p(M|D)$, in a violation of the rules of conditional probability inversion as framed by Bayes' theorem. While in well-behaved data-rich problems, the maxima of both expressions may coincide, this is an exception, not the rule. The consequences of this mistake are far-reaching as argued by [@clayton2022bernoulli], with implications extending from scientific modeling to medicine, law, and public policy.

In science, this fallacy contributes to  models fail to generalize well, an impetus to produce ad-hoc or post-hoc  uncertainty quantification, and yield published results that resist replication([@baker2016, @cobey2024biomedical]). These limiations are not confined to traditional hypothesis testing; they are embedded in the training.; they are embedded in the training and deployment of modern machine learning models as well. In regression and classification tasks, likelihood maximization is treated as sufficient for inference, even though it results in a single point estimate and fails to account for parameter uncertainty or model plausibility.

This epistemic shortcut has already been explicitly critiqued in the machine learning community. Gal [@gal2016uncertainty] and Ghahramani [@gahramani2015probabilistic] emphasize that most ML models discard uncertainty entirely and treat the output of optimization as inference. This leads to overconfident predctions and brittle generalization, echoing Clayton's concerncs. Bishop [@bishop2006pattern] also distiguishes between the utility of models and the inferential core required to quantify uncertainty, underscoring that likelihood is not enough and that the Bernoulli fallacy pervades the very logic of applied machine learning.

@@

### Limitations of Existing Approaches
Regrettably, the development of traditional ocean color predictive models relies on a fundamental statistical error that plagues most of observational science today; that of conflating sampling probability, with inferential probability([@jaynes2003probability, @DeScheemaekere2011].) Consider a dataset $D$ that consists of input-output pairs - e.g. Remote sensing reflectance (Rrs) and  chlorophyll-*a* concentration $Chl_a$ - and a model $M$, such as OCx, hypothesized to describe the statistical association between them. The sampling probability $p(D|M)$ represents the probability of observing the data $D$ if model $M$ were "true". Standard model fitting maximizes this quantity (i.e. the likelihood) by tuning the parameters of $M$.

The unspoken assumption is that the model that best fits the data also best represents the underlying process. This constitutes an epistemic fallacy; substituting $p(D|M)$ for $p(M|D)$, in a violation of the rules of conditional probability inversion as framed by Bayes' theorem. While in well-behaved data-rich problems, the maxima of both expressions may coincide, this is an exception, not the rule. The consequences of this mistake are far-reaching as argued by [@clayton2022bernoulli], with implications extending from scientific modeling to medicine, law, and public policy.

In science, this fallacy contributes to  models fail to generalize well, an impetus to produce ad-hoc or post-hoc  uncertainty quantification, and yield published results that resist replication([@baker2016, @cobey2024biomedical]). These limiations are not confined to traditional hypothesis testing; they are embedded in the training.; they are embedded in the training and deployment of modern machine learning models as well. In regression and classification tasks, likelihood maximization is treated as sufficient for inference, even though it results in a single point estimate and fails to account for parameter uncertainty or model plausibility.

This epistemic shortcut has already been explicitly critiqued in the machine learning community. Gal [@gal2016uncertainty] and Ghahramani [@gahramani2015probabilistic] emphasize that most ML models discard uncertainty entirely and treat the output of optimization as inference. This leads to overconfident predctions and brittle generalization, echoing Clayton's concerncs. Bishop [@bishop2006pattern] also distiguishes between the utility of models and the inferential core required to quantify uncertainty, underscoring that likelihood is not enough and that the Bernoulli fallacy pervades the very logic of applied machine learning.

>>>>

Satellite ocean color remote sensing has long served as a cornerstone of marine ecosystem monitoring, offering global and synoptic coverage of surface ocean properties. Among these, chlorophyll-a ($Chl_a$) concentration remains a central metric, widely used as a proxy for phytoplankton biomass, primary production, and water quality. The retrieval of $Chl_a$ from ocean color data has evolved over decades, resulting in a diverse lineage of empirical and semi-empirical algorithms. The following section summarizes this historical development, which sets the stage for a critical examination of the statistical foundations underlying current approaches.

### Historical Context of Chlorophyll Algorithms

Satellite ocean color observations have long been fundamental for monitoring marine ecosystems, as they enable global estimation of chlorophyll‑a ($Chl_a$) — a key indicator of phytoplankton biomass and ocean productivity. Early empirical algorithms, notably developed by O’Reilly et al. [@oreilly1998; @oreilly2000], established the $OCx$ family (where $x$ denotes the number of bands used) of polynomial regression models. These models relate blue-to-green reflectance ratios (after log‑transformation) to in situ $Chl_a$, employing either straight band ratios (BR) or maximum band ratios (MBR)—the latter selecting the highest available blue-to-green ratio for any given observation as input to a high‑order polynomial. These formulations have served as the operational foundation for chlorophyll‑a products across a broad range of satellite ocean color sensors—from the pioneering Coastal Zone Color Scanner (CZCS) through SeaWiFS, MODIS, and MERIS to more recent missions—offering a straightforward and robust approach for Case‑1 waters. However, their performance is more limited in optically complex Case‑2 waters and remains sensitive to atmospheric correction errors.

@@

### Limitations of Existing Approaches

Regrettably, the development of traditional ocean color algorithms is grounded in a fundamental statistical error—one that pervades much of observational science: the conflation of sampling probability with inferential probability [@jaynes2003probability; @DeScheemaekere2011].

Consider a dataset $D$ composed of input–output pairs—e.g., remote sensing reflectance (Rrs) and chlorophyll-a concentration ($Chl_a$)—and a model $M$, such as OCx, posited to represent the relationship between them. The sampling probability $p(D \mid M)$ denotes the probability of observing data $D$ under the assumption that model $M$ is true. In standard model fitting, this likelihood is maximized by adjusting the parameters of $M$ to best explain the observed data.

This approach tacitly assumes that the model which best fits the data also most accurately represents the underlying generative process. This constitutes an epistemic fallacy—treating $p(D \mid M)$ as if it were $p(M \mid D)$—a direct violation of Bayes’ theorem and the rules governing conditional probability.

Although in well-behaved, data-rich cases—where the likelihood is regular, the signal strong, and the model adequately constrained—the maxima of $p(D \mid M)$ and $p(M \mid D)$ may coincide, this remains the exception—not the rule.

This mistake lies at the heart of what Clayton [@clayton2022bernoulli] terms the Bernoulli Fallacy: the widespread misinterpretation of likelihood as inference, or of data-fit as belief. As Clayton argues, this logical misstep has far-reaching consequences, with implications that extend beyond science to domains such as medicine, law, and public policy.

In scientific modeling, this fallacy contributes to poor generalization, drives the use of ad hoc or retrospective uncertainty quantification, and underlies many published results that later prove difficult to replicate [@baker2016; @cobey2024biomedical]. These limitations are not restricted to classical hypothesis testing; they persist in the training and deployment of modern machine learning models as well.

In regression and classification, maximizing likelihood is often treated as sufficient for inference—despite yielding only a single point estimate and ignoring both parameter uncertainty and the plausibility of alternative models.

This epistemic shortcut has been directly critiqued in the machine learning literature. Gal [@gal2016uncertainty] and Ghahramani [@gahramani2015probabilistic] point out that most ML models discard uncertainty altogether, treating the outcome of an optimization as if it were an inference. The result is overconfident predictions and brittle generalization—concerns that echo Clayton’s critique.

Bishop [@bishop2006pattern] similarly distinguishes between the utility of predictive models and the inferential scaffolding required to quantify uncertainty, reinforcing the notion that likelihood alone is insufficient—and that the Bernoulli Fallacy permeates much of applied machine learning.

### Overcoming limitations

There have been attempts to address these issues. [@seegers2018] have proposed alternative metrics to circumvent the inadequate assumptions of the frequentist approach. Others have tried to go a step futher incorporate Bayesian concepts. E.g. [ @frouin2013] have proposed a Bayesian inversion scheme for atmospheric correction. [@shi2015] proposed a probabilistic method to merge data from different sensors. [@craig2019] have proposed a Bayesian neural network (BNN) approach using Hamiltonian Monte Carlo sampling to retrieve Inherent Optical Properties (IOP) from Top-of-the-atmosphere (TOA) radiance. [@werther2022] used Monte-Carlo dropout to approximate a deep BNN. [@erickson2023] have proposed using conjugate Gaussian prior and likelihood to frame the GIOP as Bayesian model to predict IOPs with uncertainty. [@hammout2024] have proposed a BNN approximation via Stochastic Variational Inference to predict $Chl_a$ from ocean color observations. Yet most of these approaches retain variable levels of frequentism by applying only part of what is now commonly referred to as the Bayesian workflow[@bgelman2019].

@@

### Limitations of Existing Approaches
Regrettably, the development of traditional ocean color algorithms is grounded in a fundamental statistical error—one that pervades much of observational science: the conflation of sampling probability with inferential probability [@jaynes2003probability; @DeScheemaekere2011].

Consider a dataset $D$ composed of input–output pairs—e.g., remote sensing reflectance (Rrs) and chlorophyll-a concentration ($Chl_a$), and a model $M$, such as OCx, posited to represent the relationship between them. The sampling probability $p(D|M)$ denotes the probability of observing data $D$ under the assumption that model $M$ is true. In standard model fitting, this likelihood is maximized by adjusting the parameters of $M$ to best explain the observed data.

This approach tacitly assumes that the model which best fits the data also most accurately represents the underlying generative process. This constitutes an epistemic fallacy—treating $p(D|M)$ as if it were $p(M | D)$; a violation of Bayes’ theorem and the rules governing conditional probability.

Although in well-behaved, data-rich cases—where the likelihood is regular, the signal strong, and the model adequately constrained—the maxima of $p(D|M)$ and $p(M|D)$ may coincide, this remains the exception rather than the rule.

This mistake lies at the heart of what Clayton [@clayton2022bernoulli] terms the Bernoulli Fallacy: the misconception that empirical probabilities by themselves yield inferential truths. As Clayton argues, this logical misstep has far-reaching consequences, with implications that extend beyond science to domains such as medicine, law, and public policy.

In scientific modeling, this fallacy contributes to poor generalization, drives the use of ad hoc or retrospective uncertainty quantification, and underlies many published results that later prove difficult to replicate [@baker2016; @cobey2024biomedical]. These limitations are not restricted to classical hypothesis testing; they persist in the training and deployment of modern machine learning models as well.

In regression and classification, maximizing likelihood is often treated as sufficient for inference—despite yielding only a single point estimate and ignoring both parameter uncertainty and the plausibility of alternative models.

This epistemic shortcut has been directly critiqued in the machine learning literature. Gal [@gal2016uncertainty] and Ghahramani [@ghahramani2015probabilistic] point out that most ML models discard uncertainty altogether, treating the outcome of an optimization as if it were an inference. The result is overconfident predictions and brittle generalization—concerns that echo Clayton’s critique.

Bishop [@bishop2006pattern] similarly distinguishes between the utility of predictive models and the inferential scaffolding required to quantify uncertainty, reinforcing the notion that likelihood alone is insufficient, and that the Bernoulli Fallacy permeates much of applied machine learning.
